Is high social class always beneficial for survival? A study of Northern Sweden 1801--2013

### 2017-10-26 article3.Rmd
Spelling correction

One of the central aspects of survival is social class and access to economic
and other resources. In present-day welfare societies, social position is a
strong determinant when it comes to health and mortality and the impact even
seems to be increasing [@kuetal04; @macketal16; @frlu07; @broba12; @strand10]. @liph95 suggest that
"... social conditions have been, are and will continue to be irreducible
determinants of health outcomes and therefore deserves their appellation as
'fundamental causes' of disease and death". The persistence of social inequality in mortality to the disadvantage of the lower classes is one of the main assumptions in their theory [@liph98]. The reigning view has been that socio-economic health and mortality inequalities were large in historical societies, probably larger than in modern societies. This is a
reasonable assumption since these societies were often characterised by
enourmous socioeconomic differences. Knowing that access to different kinds of resources
provide advantages in all aspects of life, the health advantage of higher
classes ought to be obvious.

@@

An intriguing aspect of this study is the lack of association between level of economic inequality and social differences in mortality. Even though our study do not directly address the hypothesis raised by @wipi09, our results are of relevance for this question as well. The early 20th century, in Sweden as well as in other countries, was characterised by enourmous economic differences, still this did in our regions not lead to any consistent advantage for the highest social class. After the Second World War, economic differences diminished, and was at its lowest in Sweden at about 1980, i.e. at a time when the expected pattern of a social hierarchy in survival appeared. The period thereafter is characterized by increasing economic differences, coinciding with increasing health differences as well. The long-term relation between economic inequality and health is still a puzzle requiring further studies.

>>>>

One of the central aspects of survival is social class and access to economic
and other resources. In present-day welfare societies, social position is a
strong determinant when it comes to health and mortality and the impact even
seems to be increasing [@kuetal04; @macketal16; @frlu07; @broba12; @strand10]. @liph95 suggest that
"... social conditions have been, are and will continue to be irreducible
determinants of health outcomes and therefore deserves their appellation as
'fundamental causes' of disease and death". The persistence of social inequality in mortality to the disadvantage of the lower classes is one of the main assumptions in their theory [@liph98]. The reigning view has been that socio-economic health and mortality inequalities were large in historical societies, probably larger than in modern societies. This is a
reasonable assumption since these societies were often characterised by
enormous socioeconomic differences. Knowing that access to different kinds of resources
provide advantages in all aspects of life, the health advantage of higher
classes ought to be obvious.

@@

An intriguing aspect of this study is the lack of association between level of economic inequality and social differences in mortality. Even though our study do not directly address the hypothesis raised by @wipi09, our results are of relevance for this question as well. The early 20th century, in Sweden as well as in other countries, was characterised by enormous economic differences, still this did in our regions not lead to any consistent advantage for the highest social class. After the Second World War, economic differences diminished, and was at its lowest in Sweden at about 1980, i.e. at a time when the expected pattern of a social hierarchy in survival appeared. The period thereafter is characterized by increasing economic differences, coinciding with increasing health differences as well. The long-term relation between economic inequality and health is still a puzzle requiring further studies.
